### 2024-04-03 Chapters/Chapter4.tex
Some updates after review from OPP
Uppfært eftir ábendingar frá ÓPP

\label{Chapter3} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{ChapterX}

@@

\section{Model structure}
The structure of the neural network is such that it contains some number n of  fully connected layers and batch normalization for each layer, along with regularization. All of these layers have the same number of units. The last layer has a dropout of 50\%. In addition to these layers there is one more output layer. This is simply a dense layer with 1 unit. A grid search was performed to figure out the best hyper parameters. These hyperparameters include number of units in each layer, number of epochs to train for, number of layers, batch size, optimizer and penalty to enforce in the regularization. The possible combinations tried can be seen in table \ref{table:gridSearchHyperparamters}.

\begin{table}[h]
    \centering
    \begin{tabular}{c|c}
        Layers &  4, \textcolor{red}{5}, 6, 7, 8, 9\\\hline
        Units &  128, 256, 512\\\hline
        Epochs & 100, 200, 500\\\hline
        Batch size & 32, 64, 128, 256\\\hline
        Optimizers & Adam, RMSprop, Adamax\\\hline
        Penalties & 1, 0.1, 0.01, 0\\\hline
    \end{tabular}
    \caption{Hyperparamter search with best performing combination colored red}
    \label{table:gridSearchHyperparamters}
\end{table}

\section{Feature selection and importance}
Using the three datasources, we train a model. The data from Veðurstofa only represents the ground truth and is not used as part of the training data. We use the gust ($f_g$) and wind speed ($f$) to calculate the target gust factor. From the Carra reanalysis data we get all of our weather information, with variables wind speed, wind direction, temperature, pressure. All of these variables are queried in three height levels (15, 250, 500 meters above ground) at the location of a given weather station. In addition we have the elevation above sea level over Iceland. We select the elevation in a couple of different shapes. A sector upwind, a sector upwind and downwind as well as a circle surrounding a weather station.

To add to these observed values and reanalysis, we calculate two derived variables, the Richardson number and the Brunt-Väisälä frequency. To begin with we simply use all the features we have to train our model along with the derived variables to see. We do this to be able to then use tools such as Shapley to see which features are impacting our predictions. Using Shapley values to see which features, will then allow us to exclude features that don't affect our model prediction and simplify the training data. We thus start with much higher dimensionality of our training data and reduce it as much as we can without impacting our results. This crystallizes in the landscape points. We start with n points that describe the elevation of the landscape around our weather station. This n might be in the hundreds. If we look at concentric circles outward from the weather station, we might have 10 circles (largest with radius 20 km), each with 72 points (5°spacing) and end up with 720 points for the landscape. This might slow down training. In addition, we might imagine that we are more interested in certain important landscape features. There might be a large mountain up- or downwind of our weather station, that might be the most important feature, in addition to some hills and such. To address this we use principal component analysis (PCA) to try to reduce the dimensionality of our landscape data. Another thing that might serve the same purpose is to select some number of m points that have the most increase or decrease in elevation from the weather station along with the coordinates of those points in relation to the weather station and the direction of the wind.

Using all the variables from Carra, along with the two derived factors (Ri, N) and two sectors (upwind and downwind) downscaled to 10 features using PCA we can see the feature importance in figure \ref{fig:ShapleyWaterfallFirstTry}. Looking at the waterfall graph we can see that Richardson number for higher levels is most important.

\begin{figure}[h]
    \centering
    \includegraphics[scale = 0.6]{Figures/shap_bar_nn_256_example100.png}
    \caption{Feature importance of a neural network as described with 5 hidden layers and 256 units in each}
    \label{fig:ShapleyWaterfallFirstTry}
\end{figure}

>>>>

\label{Chapter4} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{ChapterX}

@@

\section{Model structure}
The structure of the neural network is such that it contains some number $n$ of fully connected layers and batch normalization for each layer, along with regularization. All of these layers have the same number of units. The last layer has a dropout of 50\%. In addition to these layers there is one more output layer. This is simply a dense layer with 1 unit. A grid search was performed to determine the hyperparameters that minmize the loss. Hyperparameters are parameters that are set before the training begins\cite{hyperparameter_definition}. These hyperparameters include number of units in each layer, number of epochs to train for, number of layers, batch size, optimizer and penalty to enforce in the regularization. The possible combinations tried can be seen in Table \ref{table:gridSearchHyperparamters}.

\begin{table}[h]
    \centering
    \caption{Hyperparamter search with best performing combination colored red}
    \label{table:gridSearchHyperparamters}
    \begin{tabular}{c|c}
        Layers &  4, \textcolor{red}{5}, 6, 7, 8, 9\\\hline
        Units &  128, 256, 512\\\hline
        Epochs & 100, 200, 500\\\hline
        Batch size & 32, 64, 128, 256\\\hline
        Optimizers & Adam, RMSprop, Adamax\\\hline
        Penalties & 1, 0.1, 0.01, 0\\\hline
    \end{tabular}
\end{table}

\section{Feature selection and importance}
Using the three datasources, a model is trained. The data from Veðurstofa only represents the ground truth and is not used as part of the training data. The gust ($f_g$) and wind speed ($f$) are used to calculate the target gust factor. All weather parameters used in predictions come from the Carra reanalysis data, with variables wind speed, wind direction, temperature, pressure. All of these variables are queried in three height levels (15, 250, 500 meters above ground) at the location of a given weather station. The second data source for the training data, is the elevation data. Veðurstofa provided a GeoTif file that contains elevation information above sea level, covering Iceland, in a 20 by 20 grid. A couple of different landscape distributions were looked at. A sector upwind, a sector upwind and downwind as well as a circle surrounding a weather station.

To add to these observed values and reanalysis, two derived variables, the Richardson number and the Brunt-Väisälä frequency, were calculated. To begin with all the available parameters were used to train the model along with the derived variables to see. This is done to be able to then use tools such as Shapley to see which features are impacting the predictions. Using Shapley values to see which features, will then enable the exclusion of features that don't affect the model prediction and simplify the training data. This means starting with much higher dimensionality of our training data and reduce it as much as possible without impacting our results. This crystallizes in the landscape points. Starting with $n$ points that describe the elevation of the landscape around our weather station. This $n$ might be in the hundreds. Looking at concentric circles outward from a given weather station, there might be 10 circles (largest with radius 20 km), each with 72 points (5° spacing) and end up with 720 points for the landscape. This might slow down training. In addition, certain landscape features might be more important than others. There might be a large mountain up- or downwind of a given weather station, that might be the most important feature. To address this principal component analysis (PCA) is used to try to reduce the dimensionality of our landscape elevation data. PCA is a statistical procedure that allows the summarization of multivariate data to lower dimension and thus can ease visualization or increase speed of training a complex model with only minimal adverse effect to performance\cite{pca_information}. Another thing that might serve the same purpose is to select some number of $m$ points that have the most increase or decrease in elevation from the weather station along with the coordinates of those points in relation to the weather station and the direction of the wind.

Using all the variables from Carra, along with the two derived factors ($Ri$, $N$) and two sectors (upwind and downwind) downscaled to 10 features using PCA, the feature importance can be seen in Figure \ref{fig:ShapleyWaterfallFirstTry}. The waterfall graph in Figure \ref{fig:ShapleyWaterfallFirstTry} shows that the Richardson number for higher levels is most important.

\begin{figure}[h]
    \caption{Feature importance of a neural network as described with 5 hidden layers and 256 units in each}
    \label{fig:ShapleyWaterfallFirstTry}
    \centering
    \includegraphics[scale = 0.6]{Figures/shap_bar_nn_256_example100.png}
\end{figure}

### 2024-04-03 MSc.tex
Uppfært eftir ábendingar frá ÓPP

% Packages
\usepackage[utf8]{inputenc}
\usepackage[icelandic, english]{babel}
\usepackage{t1enc}
\usepackage{graphicx}
\usepackage[intoc]{nomencl}
\usepackage{enumerate,color}
\usepackage{url}
\usepackage[pdfborder={0 0 0}]{hyperref}
\BeforeTOCHead[toc]{\cleardoublepage\pdfbookmark{\contentsname}{toc}} % Add Table of Contents to PDF "bookmark" table of contents
\usepackage{appendix}
\usepackage{eso-pic}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[sf,normalsize]{subfigure}
\usepackage[format=plain,labelformat=simple,labelsep=colon]{caption}
\usepackage{placeins}
\usepackage{tabularx}
\usepackage{enumitem}
\usepackage{listings}

@@

\def\thesisyear{2024}       					% Year thesis submitted
\def\thesismonth{June}					% Month thesis submitted, e.g. "December"
\def\thesisauthor{Brynjar Geir Sigurðsson}				% Thesis authoreiningaraðferðinni
\def\thesistitle{A NN approach to predicting gust factors in complex landscape} % Title of thesis
%\def\thesissubtitle{XXSubtitle is rarely usedXX}		% Subtitle of thesis (optional)
\def\thesisshorttitle{} 	% Optional: if title of thesis is longer than 50 characters, it would not fit on the spine (kjölur) and in this case you, need to provide a short title here for the print shop. Otherwhise: make it empty
\def\thesiscredits{60 ECTS} 						% Credits awarded for the project
\def\thesissubject{Mechanical Engineering}
\def\thesiskind{M.Sc.}					% E.g. M.Sc. or Ph.D. thesis
\def\thesiskindformal{CCMagister Scientiarum}			% E.g. Magister Scientiarum
\def\thesisschool{School of Engineering and {Natural Sciences}}		% School
\def\thesisfaculty{Industrial Engineering, Mechanical Engineering and Computer Science}% Faculty name without "Faculty of" (gets added) 
\def\thesisaddress{Dunhagi 5}			        % Faculty or school office address street
\def\thesispostalcode{107}			                % Faculty or school office zip code
\def\thesistelephone{525 4000}					% Office telephone
%\def\thesispublisher{XX}					% Publisher (not used)
\def\thesissupervisors{Kristján Jónasson // Ólafur Pétur Pálsson}					% Names of supervisors (split by \\ if more than one).
\def\thesisnrofsupervisors{2}					% Number of supervisors (to use "Supervisor" vs. "Supervisors")
\def\thesiscommittee{Kristján Jónasson \\ Ólafur Pétur Pálsson}			% Thesis commitee must include the supervisor(s)
\def\thesisexaminer{XXNN3XX}				        % Examiner (prófdómari)
\def\thesisISBN{}           					% Thesis ISBN number (keep empty: not used anymore)
\def\thesisprinting{}						% Name of printsthop (keep empty if thesis is never printed)
%\def\thesisprinting{Háskólaprent, Fálkagata 2, 107 Reykjavík}	% Name of printsthop (keep empty if thesis is never printed)
\def\thesislicense{This thesis may not be copied in any form without author permission.} % Set license here (could also be some Creative Commons license)
%\def\thesiskeywords{Keyword1, Keyword2, Keyword3}		% Keywords (not used anywhere, hence commented out)

@@

>>>>

% Packages
\usepackage[utf8]{inputenc}
\usepackage[icelandic, english]{babel}
\usepackage{t1enc}
\usepackage{graphicx}
\usepackage[intoc]{nomencl}
\usepackage{enumerate,color}
\usepackage{url}
\usepackage[pdfborder={0 0 0}]{hyperref}
\BeforeTOCHead[toc]{\cleardoublepage\pdfbookmark{\contentsname}{toc}} % Add Table of Contents to PDF "bookmark" table of contents
\usepackage{appendix}
\usepackage{eso-pic}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[sf,normalsize]{subfigure}
\usepackage[format=plain,labelformat=simple,labelsep=colon]{caption}
\usepackage{placeins}
\usepackage{tabularx}
\usepackage{enumitem}
\usepackage{listings}
\usepackage{xspace}

@@

\def\thesisyear{2024}       					% Year thesis submitted
\def\thesismonth{June}					% Month thesis submitted, e.g. "December"
\def\thesisauthor{Brynjar Geir Sigurðsson}				% Thesis authoreiningaraðferðinni
\def\thesistitle{A NN approach to predicting gust factors in complex landscape} % Title of thesis
%\def\thesissubtitle{XXSubtitle is rarely usedXX}		% Subtitle of thesis (optional)
\def\thesisshorttitle{} 	% Optional: if title of thesis is longer than 50 characters, it would not fit on the spine (kjölur) and in this case you, need to provide a short title here for the print shop. Otherwhise: make it empty
\def\thesiscredits{60} 						% Credits awarded for the project
\def\thesissubject{Mechanical Engineering}
\def\thesiskind{M.Sc.}					% E.g. M.Sc. or Ph.D. thesis
\def\thesiskindformal{CCMagister Scientiarum}			% E.g. Magister Scientiarum
\def\thesisschool{School of Engineering and {Natural Sciences}}		% School
\def\thesisfaculty{Industrial Engineering, Mechanical Engineering and Computer Science}% Faculty name without "Faculty of" (gets added) 
\def\thesisaddress{Dunhagi 5}			        % Faculty or school office address street
\def\thesispostalcode{107}			                % Faculty or school office zip code
\def\thesistelephone{525 4000}					% Office telephone
%\def\thesispublisher{XX}					% Publisher (not used)
\def\thesissupervisors{Kristján Jónasson}					% Names of supervisors (split by \\ if more than one).
\def\thesisnrofsupervisors{1}					% Number of supervisors (to use "Supervisor" vs. "Supervisors")
\def\thesiscommittee{Kristján Jónasson \\ Ólafur Pétur Pálsson}			% Thesis commitee must include the supervisor(s)
\def\thesisexaminer{XXNN3XX}				        % Examiner (prófdómari)
\def\thesisISBN{}           					% Thesis ISBN number (keep empty: not used anymore)
\def\thesisprinting{}						% Name of printsthop (keep empty if thesis is never printed)
%\def\thesisprinting{Háskólaprent, Fálkagata 2, 107 Reykjavík}	% Name of printsthop (keep empty if thesis is never printed)
\def\thesislicense{This thesis may not be copied in any form without author permission.} % Set license here (could also be some Creative Commons license)
%\def\thesiskeywords{Keyword1, Keyword2, Keyword3}		% Keywords (not used anywhere, hence commented out)

@@

% To be able to use units in equations. From https://qerub.se/typesetting-units-in-latex
\newcommand{\unit}[1]{\ensuremath{\, \mathrm{#1}}}
